### 2017-04-13 paper/oopsla17-submission/abstract.tex
more intro2, plus abstract and conclusion

\begin{abstract}
  Localizing type errors is challenging in languages with global type
  inference, as the type checker must make assumptions about what the
  programmer intended to do.
  %

  We present a data-driven approach to type error localization that uses
  machine learning to assign blame based on previous
  ill-typed programs and their fixes.
  %
  First, we gather a large set of over 4,500 ill-typed student programs
  and their subsequent fixes.
  %
  Second, we frame type error localization as a supervised learning
  problem: to predict, for each expression in the
  program, whether it should be blamed.
  %
  We represent each expression by its syntactic class and type and those
  of its parent and children, associating blame with expressions that
  were changed in the human-fixed program.
  %
  Finally, we train and evaluate a series of off-the-shelf classifiers
  on our suite of programs,
  and find that they improve localization accuracy from a state-of-the-art
  56\% to 72\%.
\end{abstract}

>>>>

\begin{abstract}
Localizing type errors is challenging in
languages with global type inference, as
the type checker must make assumptions
about what the programmer intended to do.
%
We introduce \toolname, a \emph{data-driven}
approach to error localization based on
supervised learning.
%
\toolname works by using a large corpus
of training data --- pairs of ill-typed
programs and their ``fixed'' versions ---
to automatically \emph{learn a model}
of where the error is most likely
to be found.
%
Given a new ill-typed program,
\toolname executes the model to
generate a list of potential blame
assignments ranked by likelihood.
%
We evaluate \toolname by comparing its
precision to the state of the art
on a set of over 4,500 ill-typed Ocaml
programs drawn from introductory
programming classes.
%
We show that when the top-ranked blame assignment
is considered, \toolname's data-driven
model is able to correctly predict
the exact sub-expression that should
be changed 74\% of the time, %which is
29 points higher than \ocaml and
18 points higher than the state-of-the-art
\sherrloc tool.
%
Furthermore, \toolname's accuracy surpasses
85\% when we consider the top \emph{two}
locations and reaches 91\% if consider
the top \emph{three}.
\end{abstract}

### 2017-04-13 paper/oopsla17-submission/conclusion.tex
more intro2, plus abstract and conclusion

We have presented a new data-driven approach to localizing type errors,
using machine learning to assign blame based on previously seen novice
mistakes.
%
We translate ill-typed programs into sets of feature vectors, predict
for each sub-expression whether it should be blamed, rank the
predictions based on the model's confidence, and select the top three
blame assignments to present to the user.
%
We represent expressions solely by their syntactic class and type, and
those of their parent and children.
%
We have evaluated our approach with five different learning algorithms
on a suite of over 4,500 student programs, and found that all of our
models outperform the state of the art, by up to a 16\% margin.
%

>>>>

We have presented \toolname, which
combines modern statistical methods
with domain-specific feature engineering
to open the door to a new data-driven
path to precise error localization.
%
\toolname
translates pairs of
blame-labeled ill-typed
programs into bags-of-abstracted-terms,
and then trains a classifier
to predict whether a feature vector
should get a blame label.
% that
% predicts the value of the
% whether a feature vector gets
% a blame label.
%
Given a new ill-typed program,
we extract its bag-of-extracted-terms
to compute the blame label for each
term and rank the predictions based
on the model's confidence, to select
the top 1-3 blame assignments to
present to the user.

We have evaluated our approach with
five different classifiers on a suite
of over 4,500 student programs.
%
We found that while machine learning
over syntactic features of each term in isolation
performs worse than existing
purely constraint based approaches, % (\eg \ocaml, \sherrloc),
augmenting the data with a single feature corresponding to
the type error slice brings our
classifiers up to par with the state of the art,
and further augmenting the data with
features of an expression's parent and children
allows our classifiers to outperform
the state of the art by 18 percentage points.

% We find that while machine
% learning on its own performs slightly worse
% than existing purely constraint based
% approaches (\eg \ocaml, \sherrloc), once
% we augment the data with feature abstraction
% corresponding to the \emph{type error slice}~\cite{Tip2001-qp},
% the resulting models models outperform
% the state of the art by up to 18
% percentage points.

As with other forms of machine learning,
a key concern is that of \emph{data-set bias}: are
the models learned by \toolname specific
to our data set, and would they fail on
\emph{other} \ocaml programs?
%
We address this concern in two ways.
%
First, we partition the data by year,
and show that models learned from one
year generalize to, \ie perform nearly
as well on, the programs from the other
year.
%
Second, we argue that in our setting
this bias is a \emph{feature} (and not
a bug): it allows \toolname to \emph{adapt}
to the kinds of errors that programmers
(specifically, novices who are in greatest
need of precise feedback) actually make
rather than hardwiring the biases of
compiler authors who, by dint of their
training and experience, may suffer from
blind spots with regards to such problems.
%
In this regard, we are particularly pleased
that our classifiers can be trained on a
modest amount of data, \ie a single course's
worth, and envision a future where each course
comes equipped with a model of its students' errors.

### 2017-04-13 paper/oopsla17-submission/intro2.tex
more intro2, plus abstract and conclusion

\mypara{A Data-Driven Approach}
%
In this paper, we introduce \toolname
\footnote{``Numeric Analysis of Type Errors''; any resemblance to persons living or dead is purely coincidental.}
a \emph{data-driven} approach to error
localization based on supervised
learning (see \citealt{Kotsiantis2007-pj} for a survey).
%
\toolname works by analyzing a large corpus
of training data --- pairs of ill-typed
programs and their subsequent fixes ---
to automatically \emph{learn a model}
of where errors are most likely to
be found.
%
Given a new ill-typed program,
\toolname simply executes the model
to generate a list of potential error
locations ranked by likelihood.
%
We evaluate \toolname by comparing its
precision against the state-of-the-art
on a set of over 4,500 ill-typed \ocaml
programs drawn from introductory
programming classes.
%
We show that \toolname's data-driven
model is able to correctly predict
the exact sub-expression that needs to
be changed 74\% of the time,
29 points higher than \ocaml and
18 points higher than the state-of-the-art
\sherrloc tool.
%
Further, \toolname's accuracy surpasses
85\% when we consider the \emph{top two}
locations and reaches 91\% if consider
the \emph{top three}.
%
We achieve these advances by identifying
and then solving three key challenges.

@@

Our \emph{first contribution} (\autoref{sec:overview})
is a set of more than 4,500 labeled programs,
gives us an accurate ground truth of
the kinds of errors and the (locations
of the) fixes that novices make in
practice.
%
We obtain this set by observing that
software development is an iterative
process; programmers eventually
fix their own ill-typed programs,
perhaps after multiple incorrect
exploratory attempts.
%
To exploit this observation we instrumented
the \ocaml compiler to collect fine-grained
traces of student interactions over two instances
of an undergraduate Programming Languages
course.\footnote{
% \begin{anonsuppress}
% UC San Diego (IRB \#140608).
% \end{anonsuppress}
% \begin{noanonsuppress}
at AUTHOR's INSTITUTION (IRB HIDDEN).
% \end{noanonsuppress}
}
%
We then post-process the resulting time-series
of programs submitted to the \ocaml compiler into
a set of pairs of ill-typed programs and their
subsequent \emph{fixes}, the first (type-)~correct
program in the trace suffix.
%
Finally, we label the changed locations by computing a
\emph{tree-diff} between the two terms to find
the exact sub-terms that were changed in the fix.

@@

Our \emph{second contribution} (\autoref{sec:learning})
solves this problem with a simple, yet expressive, representation called
a \emph{Bag-of-Abstract-Terms (BOAT)} wherein
%
(1)~each program is represented by the \emph{bag}
    or multiset of (sub-)~terms that appears inside
    it; and
%
(2)~each (sub-)~term $t$ is \emph{abstracted} as
    a vector $[f_1(t), \ldots, f_n(t)]$, where
    each $f_i$ is a \emph{feature abstraction}
    that maps a term to a numeric value.
%
We can model the error labels with a simple feature
abstraction (\eg does-change-in-fix).
%
We can even account for \emph{contextual} information like
the features of the parent- and child- terms by
concatenating the feature vectors of each term with those
of its parent and children.
%
We have found this representation to be particularly
convenient as it gives us flexibility
in modeling the syntactic and semantic structure of programs
while retaining compatibility with off-the-shelf classifiers,
in contrast to, \eg, \citet{Raychev2015-jg}, who had to develop
their own variants of classifiers to obtain their results.

\mypara{Challenge 3: Training Precise Classifiers}
%
Finally, the last and most important challenge is to
use the labeled BOAT data to train classifiers that
are capable of \emph{precisely} pinpointing the error
locations in real programs.
%
The key here is to find the right set of feature
abstractions and classification algorithms, that
lead to precise predictions.
%
Fortunately, our BOAT model allows us a great deal of
latitude in our choice of features.
%
We can use abstraction functions to capture different
aspects of a term ranging from
%
syntactic features (\eg is-a-data constructor, is-a-literal,
is-an-arithmetic-operation, is-a-function-application, \etc),
%
to semantic features captured by the type system (\eg is-a-list,
is-an-integer, is-a-function, \etc).

>>>>

\mypara{A Data-Driven Approach}
%
In this paper, we introduce \toolname
\footnote{``Numeric Analysis of Type Errors''; any resemblance to persons living or dead is purely coincidental.}
a \emph{data-driven} approach to error
localization based on supervised
learning (see \citealt{Kotsiantis2007-pj} for a survey).
%
\toolname works by analyzing a large corpus
of training data --- pairs of ill-typed
programs and their subsequent fixes ---
to automatically \emph{learn a model}
of where errors are most likely to
be found.
%
Given a new ill-typed program,
\toolname simply executes the model
to generate a list of potential
blame assignments ranked by likelihood.
%
We evaluate \toolname by comparing its
precision against the state-of-the-art
on a set of over 4,500 ill-typed \ocaml
programs drawn from introductory
programming classes.
%
We show that, when restricted to a
\emph{single} prediction, \toolname's data-driven
model is able to correctly predict
the exact sub-expression that should
be changed 74\% of the time,
29 points higher than \ocaml and
18 points higher than the state-of-the-art
\sherrloc tool.
%
Furthermore, \toolname's accuracy surpasses
85\% when we consider the \emph{top two}
locations and reaches 91\% if consider
the \emph{top three}.
%
We achieve these advances by identifying
and then solving three key challenges.

@@

Our \emph{first contribution} (\autoref{sec:overview})
is a set of more than 4,500 labeled programs,
giving us an accurate ground truth of
the kinds of errors and the (locations
of the) fixes that novices make in
practice.
%
We obtain this set by observing that
software development is an iterative
process; programmers eventually
fix their own ill-typed programs,
perhaps after multiple incorrect
exploratory attempts.
%
To exploit this observation we instrumented
the \ocaml compiler to collect fine-grained
traces of student interactions over two instances
of an undergraduate Programming Languages
course.\footnote{
% \begin{anonsuppress}
% UC San Diego (IRB \#140608).
% \end{anonsuppress}
% \begin{noanonsuppress}
at AUTHOR's INSTITUTION (IRB HIDDEN).
% \end{noanonsuppress}
}
%
We then post-process the resulting time-series
of programs submitted to the \ocaml compiler into
a set of pairs of ill-typed programs and their
subsequent \emph{fixes}, the first (type-)~correct
program in the trace suffix.
%
Finally, we compute the blame labels using a
\emph{tree-diff} between the two terms to find
the exact sub-terms that were changed in the fix.

@@

Our \emph{second contribution} (\autoref{sec:learning})
solves this problem with a simple, yet expressive, representation called
a \emph{Bag-of-Abstracted-Terms (BOAT)} wherein
%
(1)~each program is represented by the \emph{bag}
    or multiset of (sub-)~terms that appears inside
    it; and
%
(2)~each (sub-)~term $t$ is \emph{abstracted} as
    a vector $[f_1(t), \ldots, f_n(t)]$, where
    each $f_i$ is a \emph{feature abstraction}
    that maps a term to a numeric value.
%
We can model the blame labels with a simple feature
abstraction (\eg does-change-in-fix).
%
We can even account for \emph{contextual} information like
the features of the parent- and child- terms by
concatenating the feature vectors of each term with those
of its parent and children.
%
We have found this representation to be particularly
convenient as it gives us flexibility
in modeling the syntactic and semantic structure of programs
while retaining compatibility with off-the-shelf classifiers,
in contrast to, \eg, \citet{Raychev2015-jg}, who had to develop
their own variants of classifiers to obtain their results.

\mypara{Challenge 3: Training Precise Classifiers}
%
Finally, the last and most important challenge is to
use the labeled BOAT data to train classifiers that
are capable of \emph{precisely} pinpointing the errors
in real programs.
%
The key here is to find the right set of feature
abstractions and classification algorithms, that
lead to precise predictions.
%
Fortunately, our BOAT model allows us a great deal of
latitude in our choice of features.
%
We can use abstraction functions to capture different
aspects of a term ranging from
%
syntactic features (\eg is-a-data constructor, is-a-literal,
is-an-arithmetic-operation, is-a-function-application, \etc),
%
to semantic features captured by the type system (\eg is-a-list,
is-an-integer, is-a-function, \etc).
